---
title: "Gov 50 Final Project"
author: "Julia English"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

In education, there are conversations surrounding what contributes most to student success. NYC department of education has open testing data, teacher survey data and demographic data at the school level that I would like to investigate particularly in relation to math scores. 


## Proposal

What affects student success in math? For this data analysis, I'm interested in working with multiple data sets from the New York City Department of Education including the 2017 School Math results, the school level demographic snapshot and the public teacher data files. There are many items that could affect student outcomes. I want to specifically look at the impact of teacher expertise on math outcomes while holding for effectiveness of administration as well as the poverty rate of the school. My outcome variable would be 2017 math scores on the school level. There are a range of possible levels, but possible scores range from 134-400 My predictor variable would be teacher effectiveness scores measured on a scale from 1-10. I want to run a multiple regression as variables in education are usually intertwined. For example, teaching a high percentage of English Language Learners might affect how students complete complicate state test word problems. Holding these covariants constant and still seeing a statistically significant association would be convincing evidence. 


```{r}
library(tidyverse)
library(ggplot2)
library(foreign)

teachers <- read.csv("2017_Public_Data_File_Teacher.csv")
math <- read.csv("2013-2017_School_Math_Results_-_All.csv")
demographics <- read.csv("2013_-_2018_Demographic_Snapshot_School.csv")
```

```{r univariate distributions}



math_sub <- math %>% 
  filter(Year == 2017,
         Grade == 8) %>% 
  select(DBN, School.Name, Year, Number.Tested, Mean.Scale.Score) 

math_sub$Mean.Scale.Score <- as.numeric(math_sub$Mean.Scale.Score)
  

math.plot <- math_sub %>% 
  ggplot(mapping = aes(x = Mean.Scale.Score)) +
  geom_histogram(color = "black", fill = "seagreen") +
  labs(title = "Histogram of 2017 Math Scores for the 8th Grade",
       x = "Average School Math Score",
       y = "Count") +
  geom_vline(aes(xintercept = 322), color = "violet")
  

math.plot

math_sub %>% 
  select(Mean.Scale.Score) %>% 
  summary()

```
The purple line indicates the proficiency level for 2017 for 8th grade math students. 
